openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-07-12 paper/policar2020-opentsne.tex
Add references to introduction.

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis (PCA),
multidimensional scaling, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received
much attention as it can address high volumes of data and reveal meaningful
clustering structure. Most of the recent reports on single-cell gene expression
data start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
embeddings.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence to suggest that
any data set can accurately be described using only two dimensions and that any
embedding onto a two-dimensional plane will inevitably lead to loss of
information. Therefore, it is beneficial to examine multiple embeddings, each
of which provides a different perspective into the topology and other
characteristics of the data.
\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use openTSNE to create four different
    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
    each providing a different perspective into the topology of the data.
    The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
the same gene-expression data containing cells from the mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
Warm colors correspond to excitatory neurons, while
cool colors represent inhibitory neurons. While different clusters of
excitatory and inhibitory neurons appear close to one another, all clusters
appear somewhat equidistant from their neighbors, and the overall
relations between groups are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white-space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups and may be
clustered further into cell sub-types.

>>>>

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis (PCA),
multidimensional scaling, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. 
Among these, t-SNE lately received much attention as it can address high
volumes of data and reveal meaningful clustering structure. Increased
resolution and throughput of modern molecular assays has lead to frequent use
of t-SNE in diverse fields including, but not limited to, single-cell
transcriptomics~\cite{macosko2015highly,cao2019single,tasic2018shared}, human
genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, spatial organization of microbial
communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}.

Reports on single cell gene expression data, our running example,
often start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
embeddings.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence to suggest that
any data set can accurately be described using only two dimensions and that any
embedding onto a two-dimensional plane will inevitably lead to loss of
information. Therefore, it is beneficial to examine multiple embeddings, each
of which provides a different perspective into the topology and other
characteristics of the data.

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use openTSNE to create four different
    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
    each providing a different perspective into the topology of the data.
    The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
the same gene-expression data containing cells from the mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
Warm colors correspond to excitatory neurons, while
cool colors represent inhibitory neurons. While different clusters of
excitatory and inhibitory neurons appear close to one another, all clusters
appear somewhat equidistant from their neighbors, and the overall
relations between groups are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white-space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups and may be
clustered further into cell sub-types.
